---
title: "Prompts 2.0 analyses"
author: "Sophia Minnillo"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '4'
  html_notebook:
    df_print: paged
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: '4'
    df_print: paged
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

```{r, include=FALSE, echo=FALSE}
## SET UP ####
#install.packages("tidyverse")
library(tidyverse)
library(Rmisc) #for SummarySE
library(lme4)
library(lmerTest)
library("MASS")
library(MuMIn)
library(emmeans)
library(nlme) #for designing models that deal with autocorrelation
library(car) #for tests-- assumptions
library(multilevelTools) #for assumptions
library(JWileymisc)
library(effectsize)
#install.packages("remotes")  # if not installed
#remotes::install_github("glmmTMB/glmmTMB/glmmTMB")
library(glmmTMB) #special type of GLMM for right skewed data
library(DHARMa) #model fit
library(performance) #checks model performance
library(effectsize)

#load csvs
lexis_012726 <- read_csv('TAALES_results_020526.csv')
#view(lexis_012726)

#past tense csv
tense_020526 <- read_csv('prompts2_pret_imp_020526_5.csv')
#view(tense_020526)
#updated 2-19-26 based on Andrea manually cleaning the spacy output
```

```{r, echo=FALSE, include=FALSE}
#export list of students who didn't produce any past tenses
# metadata_prompts2 <- lexis_012726|>
#   dplyr::select(c(filename, Participant, Course, Prompt))
# #view(metadata_prompts2)
# #view(tense_020526_sum)
# 
# #38 students, 76 texts
# 
# #left join tense csv by metadata. Need to use tense_020526_sum from later!
# metadata_prompts2_tense <- metadata_prompts2|>
#   left_join(tense_020526_sum, by = c('Participant', 'Course', 'Prompt'))
#view(metadata_prompts2_tense)
#write_csv(metadata_prompts2_tense, 'prompts_who_produced_past_020626.csv')

#76

#how many cases that students didn't produce any past tense: 14 texts
```


# Lexis Analysis

## Inspect Data

```{r, echo=FALSE}
hist(lexis_012726$MATTR_lemmas)
#fairly normally distributed

hist(lexis_012726$lemma_frequency_log_CW)
#a bit of a right skew-- more high frequency words
#could try to keep transforming

hist(lexis_012726$lexical_density)
#steep curve, although averages out in center
```

## Graph Data

```{r, echo=FALSE}
#MAATR
ggplot(data = lexis_012726,
       aes(x = Prompt, y = MATTR_lemmas, color = Prompt)) +
  geom_violin(draw_quantiles = .50) + 
  #this shows the distribution of scores more clearly than a box plot
  geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(x = "Prompt", y = "Lexical Diversity")+
  ylim(0,1)+
  scale_color_brewer(palette="Dark2")

#lexical sophistication
ggplot(data = lexis_012726,
       aes(x = Prompt, y = lemma_frequency_log_CW, color = Prompt)) +
  geom_violin(draw_quantiles = .50) + 
  #this shows the distribution of scores more clearly than a box plot
  geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(x = "Prompt", y = "Lexical Sophistication")+
  ylim(0,13)+
  scale_color_brewer(palette="Dark2")

#lexical density
ggplot(data = lexis_012726,
       aes(x = Prompt, y = lexical_density, color = Prompt)) +
  geom_violin(draw_quantiles = .50) + 
  #this shows the distribution of scores more clearly than a box plot
  geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(x = "Prompt", y = "Lexical Density")+
  ylim(0,1)+
  scale_color_brewer(palette="Dark2")
```

## Run LMER models

Remember, Bonferroni correction: 0.00714 - statistical significance threshold.

### MATTR
```{r, echo=FALSE}
#1a. How well do the prompts predict learners' lexical diversity?

#mod1: no random slopes, random intercept by participant
mod1 <- lmer(MATTR_lemmas ~ Prompt + (1|Participant), 
             lexis_012726,  
             control=lmerControl(optimizer="bobyqa",
                                 optCtrl=list(maxfun=2e5)))
summary(mod1)
r.squaredGLMM(mod1)
emmeans(mod1, specs = pairwise~Prompt)


#effect size

#first have to pivot tibble wider
lexis_012726_cd <- lexis_012726|>
  dplyr::select(Participant, Prompt, MATTR_lemmas)|>
  pivot_wider(names_from = Prompt, values_from = MATTR_lemmas)
#view(lexis_012726_cd)

#Cohen's d
effectsize::cohens_d(lexis_012726_cd$bad_time, lexis_012726_cd$good_time, pooled_sd = TRUE, paired = TRUE)
#"Most of the time the actual population standard deviation is not known, this is why we estimate it using a pooled standard deviation from our two groups." (National University Stats Center)


#effect size with emmeans package: ignoring https://rdrr.io/cran/emmeans/man/eff_size.html
emm <- emmeans(mod1, ~ Prompt)
eff_size(emm, sigma = sigma(mod1), edf = df.residual(mod1))

#mod2: random slopes
# mod2 <- lmer(MATTR_lemmas ~ Prompt + (1+Prompt||Participant), 
#              lexis_012726,  
#              control=lmerControl(optimizer="bobyqa",
#                                  optCtrl=list(maxfun=2e5)))
# summary(mod2)
# r.squaredGLMM(mod2)
# emmeans(mod2, specs = pairwise~Prompt)
#not working

#mod3: no random slopes, random intercept by participant & course
#singularity warning!
# mod3 <- lmer(MATTR_lemmas ~ Prompt + (1|Participant)+ (1|Course), 
#              lexis_012726,  
#              control=lmerControl(optimizer="bobyqa",
#                                  optCtrl=list(maxfun=2e5)))
# summary(mod3)
# r.squaredGLMM(mod3)
# emmeans(mod3, specs = pairwise~Prompt)
```

#### Check assumptions
```{r, echo=FALSE}
# 1. Linearity
# 2. Normality of the residuals
# 3. Homogeneity of residual variance (homoscedasticity)
# 4. No autoccorelation and no multicolinearity
# 
# more about checking assumptions here: https://ademos.people.uic.edu/Chapter18.html
# https://bookdown.org/animestina/phd_july_19/testing-the-assumptions.html
# https://library.virginia.edu/data/articles/diagnostic-plots 
# https://chenzixu.rbind.io/slides/lme/lmer#16
#https://cran.r-project.org/web/packages/multilevelTools/vignettes/lmer-vignette.html

#checking model assumptions

#1. Checking Linearity Assumption:

#let's try with sjplot
sjPlot::plot_model(mod1, type='slope')

plot(fitted(mod1),residuals(mod1))

#2. Normality of the residuals
lattice::qqmath(mod1)
qqnorm(residuals(mod1))

#3. Homogeneity of residual variance (homoscedasticity)
#let's try with sjplot
sjPlot::plot_model(mod1, type='diag')

#4. No autoccorelation and no multicolinearity
#checking autocorrelation
acf(residuals(mod1))

# Extract residuals and run durbinWatsonTest for autocorrelation
mod1_residuals <- residuals(mod1)
durbinWatsonTest(mod1_residuals)
#because result is > 2, we're good

#checking multicolinearity
performance::check_collinearity(mod1)

#JWileymisc::modelDiagnostics(mod1)
md <- modelDiagnostics(mod1, ev.perc = .001)
plot(md, ask = FALSE, ncol = 2, nrow = 3)
```

### Lexical sophistication
```{r, echo=FALSE}
#### lemma_frequency_log_CW models ####
#1b. How well do the prompts predict learners' lexical sophistication?

#mod_soph1: no random slopes, random intercept by participant
mod_soph1 <- lmer(lemma_frequency_log_CW ~ Prompt + (1|Participant), 
                  lexis_012726,  
                  control=lmerControl(optimizer="bobyqa",
                                      optCtrl=list(maxfun=2e5)))
summary(mod_soph1)
r.squaredGLMM(mod_soph1)
emmeans(mod_soph1, specs = pairwise~Prompt)
#effect size
cohens_d(lemma_frequency_log_CW ~ Prompt, data = lexis_012726, pooled_sd = TRUE)

#effect size with emmeans package
emm <- emmeans(mod_soph1, "Prompt")
eff_size(emm, sigma = sigma(mod_soph1), edf = df.residual(mod_soph1))

#No significant difference for lexical sophistication:
#high lexical sophistication anyways because speaking task
```

#### Check assumptions
```{r, echo=FALSE}
##### Check for Assumptions ####
# 1. Linearity
# 2. Normality of the residuals
# 3. Homogeneity of residual variance (homoscedasticity)
# 4. No autoccorelation and no multicolinearity
# 
# more about checking assumptions here: https://ademos.people.uic.edu/Chapter18.html
# https://bookdown.org/animestina/phd_july_19/testing-the-assumptions.html
# https://library.virginia.edu/data/articles/diagnostic-plots 
# https://chenzixu.rbind.io/slides/lme/lmer#16
#https://cran.r-project.org/web/packages/multilevelTools/vignettes/lmer-vignette.html

#checking model assumptions

#1. Checking Linearity Assumption:

#let's try with sjplot
sjPlot::plot_model(mod_soph1, type='slope')

plot(fitted(mod_soph1),residuals(mod_soph1))

#2. Normality of the residuals
lattice::qqmath(mod_soph1)
qqnorm(residuals(mod_soph1))

#3. Homogeneity of residual variance (homoscedasticity)
#let's try with sjplot
sjPlot::plot_model(mod_soph1, type='diag')

#4. No autoccorelation and no multicolinearity
#checking autocorrelation
acf(residuals(mod_soph1))

# Extract residuals and run durbinWatsonTest for autocorrelation
mod_soph1_residuals <- residuals(mod_soph1)
durbinWatsonTest(mod_soph1_residuals)
#because result is > 2, we're good

#checking multicolinearity
performance::check_collinearity(mod_soph1)

#JWileymisc::modelDiagnostics(mod_soph1)
md <- modelDiagnostics(mod_soph1, ev.perc = .001)
plot(md, ask = FALSE, ncol = 2, nrow = 3)
```

### Lexical density
```{r, echo=FALSE}
#1a. How well do the prompts predict learners' lexical density?

#mod_density1: no random slopes, random intercept by participant
mod_density1 <- lmer(lexical_density ~ Prompt + (1|Participant), 
                     lexis_012726,  
                     control=lmerControl(optimizer="bobyqa",
                                         optCtrl=list(maxfun=2e5)))
summary(mod_density1)
r.squaredGLMM(mod_density1)
emmeans(mod_density1, specs = pairwise~Prompt)

#effect size

#first have to pivot tibble wider
lexis_012726_cdLD <- lexis_012726|>
  dplyr::select(Participant, Prompt, lexical_density)|>
  pivot_wider(names_from = Prompt, values_from = lexical_density)
#view(lexis_012726_cdLD)

#Cohen's d
effectsize::cohens_d(lexis_012726_cdLD$bad_time, lexis_012726_cdLD$good_time, pooled_sd = TRUE, paired = TRUE)
#"Most of the time the actual population standard deviation is not known, this is why we estimate it using a pooled standard deviation from our two groups." (National University Stats Center)

#effect size with emmeans package
emm <- emmeans(mod_density1, "Prompt")
eff_size(emm, sigma = sigma(mod_density1), edf = df.residual(mod_density1))
```

#### Check assumptions
```{r, echo=FALSE}
# 1. Linearity
# 2. Normality of the residuals
# 3. Homogeneity of residual variance (homoscedasticity)
# 4. No autoccorelation and no multicolinearity
# 
# more about checking assumptions here: https://ademos.people.uic.edu/Chapter18.html
# https://bookdown.org/animestina/phd_july_19/testing-the-assumptions.html
# https://library.virginia.edu/data/articles/diagnostic-plots 
# https://chenzixu.rbind.io/slides/lme/lmer#16
#https://cran.r-project.org/web/packages/multilevelTools/vignettes/lmer-vignette.html

#checking model assumptions

#1. Checking Linearity Assumption:

#let's try with sjplot
sjPlot::plot_model(mod_density1, type='slope')

plot(fitted(mod_density1),residuals(mod_density1))

#2. Normality of the residuals
lattice::qqmath(mod_density1)
qqnorm(residuals(mod_density1))

#3. Homogeneity of residual variance (homoscedasticity)
#let's try with sjplot
sjPlot::plot_model(mod_density1, type='diag')

#4. No autoccorelation and no multicolinearity
#checking autocorrelation
acf(residuals(mod_density1))

# Extract residuals and run durbinWatsonTest for autocorrelation
mod_density1_residuals <- residuals(mod_density1)
durbinWatsonTest(mod_density1_residuals)
#because result is > 2, we're good

#checking multicolinearity
performance::check_collinearity(mod_density1)

#JWileymisc::modelDiagnostics(mod_density1)
md <- modelDiagnostics(mod_density1, ev.perc = .001)
plot(md, ask = FALSE, ncol = 2, nrow = 3)
```

## Qualitative: Lexical sophistication analysis

1. using spacy, extract all content words from texts (tokens and lemma forms), keeping metadata from texts
2. calculate log frequency of lemmas
3. graph distribution of lexical sophistication (violin plots)
4. demonstrate what this looks like in real student texts

```{r, echo=FALSE}

```

### examples of contrasts between the same student

P5_SPA3_bad_time_solo_palabras

Una vez estábamos en el lugar de trabajo en un restaurante y un hombre Un hombre estaba muy grave Y creo que El era Creo que el hablaba con los de Y muchos chicos y y madres Visitan a mi restaurantes Cuando un hombre que no se come nada por el restaurante No sé que quien Y mi manager tenía que Y Mi manager es mi amigo también Y un poco bajo un hombre que que no conozco Y era un poco mal pero llamé la seguridad Y otro hombre a venga y S  Pues si si Al final no pasó nada Todo estaba bien Sí Sí [Tu amigo estaba bien] Sí si si sí Pero Muchos gritos muchos ruidos En esos momentos tenía que estar listo Lista Yo se poco Un poco  Luchador no no No necesito pelear a nadie En serio puedo hacer poco pero No peleaba con nadie En serio Él es muy serio y muy tranquilo Y también sabe que pelear El boxing Pero Pero es Yo sé que El No no A él no le gusta Quiero pelear Nadie En serio

P5_SPA3_good_time_solo_palabras

La última fin de semana mi amigo tenía un fiesta por Halloween Un poco tarde pero nosotros lle llevábamos costumbres y  Y Y había muchos comida o muchas comidas y bebidas y también tra trabajó con eso amigo muchos de mi O compañeros de trabajo es es tu Estuvieron Sí Sí y era divertido Especialmente con esos personas Porque en trabajo no podemos hablar mucho porque estamos trabajando Pero A mi me gustan esos personas Esas personas A mis costumbre Era un como un cowboy Y Y mi novio era un samurai Yo comí Mi compañero de trabajo trajo sopa Un poco extraño traer sopa a una fiesta, pero a ella cono cono la cono cono Conocer con cocina Cociné cociné y es Era muy delicioso rico

P6_SPA22_bad_time_solo_palabras

Hay una vez cuando mi amiga y yo fuimos a una fiesta de de cumple cumpleaños y Cuando la fiesta termina mi amiga y yo necesitamos mi amiga y yo tenemos que necesitar caminar a mi casa y mi casa Fue cerca de la fiesta pero fue en la noche y mi amiga y yo caminamos Camina Caminamos A mi casa pero fue en la noche y hay no no Otras personas con con nos y mi casa Y la fiesta fue cercano de Pero Pero esta noche siéntete No a esta noche El camino fue muy muy largo y cuando mi amiga y yo Y llegamos a nuestras casas Estamos muy alegre y No tengo miedo Después de el camino Pero durante el camino tenemos miedo

P6_SPA22_good_time_solo_palabras

Estoy alegre Cuando está el Navidad Y en el año pasado En la Navidad Estoy con mi familia y Tuvimos a una fiesta de Navidad Hay muchos muchos regalos y muchas comidas Y mi familia y yo Pasamos una día con Pasamos una dia en la cocina y tenemos que cocinar todos la comida para la fiesta y para la Navidad Nosotros cocinamos a los galletos El carne las papas fritas y cocinamos Cocinamos otros Otros postres Tengo. Tuve mucho divertido porque Porque me gusta cocinar y me gusta mi familia Y estoy con mi familia Y la fiesta fue muy divertido pienso Pienso Recibe Los Para el esqui Para esquiar Pienso Pienso Celebramos Celebramos En el día antes de Navidad Celebramos en la casa


# Tense-Aspect Analysis

Steps:

1. Calculate number of preterit and imperfect uses per file

2. inspect data

3. run models with this data --> logistic mixed-effects regression


```{r, include=FALSE, echo=FALSE}
tense_020526_sum <- tense_020526 |>
  dplyr::group_by(Participant, Course, Prompt, Tense)|>
  dplyr::summarize(count = n())|>
  pivot_wider(names_from = Tense, values_from = count)|>
  mutate(
    preterit = replace_na(preterit, 0),
    imperfect = replace_na(imperfect, 0)
  )|>
  #join with lexis spreadsheet to get wordcount info for text, then calculate proportion
  full_join(lexis_012726)|>
  dplyr::select(c(Participant, Course, Prompt, imperfect, preterit, nwords))|>
  mutate(
    preterit = replace_na(preterit, 0),
    imperfect = replace_na(imperfect, 0)
  )|>
  dplyr::rowwise()|>
  dplyr::mutate(
    preterit_per_100_words = (preterit / nwords * 100),
    imperfect_per_100_words = (imperfect / nwords * 100)
  )

#view(tense_020526_sum)

#how many students produced absolutely no past
#12 total students, 14 cases (P44 and P30 produced no past in both prompt conditions)
```

Students with absolutely no past production: 12 total students, 14 cases (P44 and P30 produced no past in both prompt conditions)

## Inspect Data

```{r, echo=FALSE}
hist(tense_020526_sum$preterit_per_100_words)

hist(tense_020526_sum$imperfect_per_100_words)
#super right skewed! lots of 0s
```

## Graph Data

```{r, echo=FALSE}
#preterit
ggplot(data = tense_020526_sum,
       aes(x = Prompt, y = preterit_per_100_words, color = Prompt)) +
  geom_violin(draw_quantiles = .50) + 
  #this shows the distribution of scores more clearly than a box plot
  geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(x = "Prompt", y = "Preterit Use per 100 Words")+
  ylim(-5,15)+
  scale_color_brewer(palette="Dark2")

#imperfect
ggplot(data = tense_020526_sum,
       aes(x = Prompt, y = imperfect_per_100_words, color = Prompt)) +
  geom_violin(draw_quantiles = .50) + 
  #this shows the distribution of scores more clearly than a box plot
  geom_jitter(shape=16, position=position_jitter(0.2))+
  labs(x = "Prompt", y = "Imperfect Use per 100 Words")+
  ylim(-5,15)+
  scale_color_brewer(palette="Dark2")
```

## Run Tweedie mixed effects model (GLMM)

### Preterit

```{r, echo=FALSE}
#view(tense_020526_sum)
M1preterit <- glmmTMB(preterit_per_100_words ~ Prompt + (1|Participant),
            	data = tense_020526_sum,
            	family = tweedie(link = "log"))
summary(M1preterit)
r.squaredGLMM(M1preterit)
#emmeans
emm_pret <- emmeans(M1preterit, ~ Prompt, type = "response")
emm_pret
contrast(emm_pret, method = "revpairwise")  # differences on response scale

#effect size with emmeans package
emm <- emmeans(M1preterit, "Prompt")
eff_size(emm, sigma = sigma(M1preterit), edf = df.residual(M1preterit))
```

#### Check assumptions

```{r}
# Check Simulated residuals- residual diagnostics
sim_res <- DHARMa::simulateResiduals(M1preterit)
plot(sim_res)
#additional tests
DHARMa::testUniformity(sim_res)     # global model fit
DHARMa::testDispersion(sim_res)     # over/under-dispersion
DHARMa::testZeroInflation(sim_res)  # unexpected excess zeros
```

### Imperfect

```{r, echo=FALSE}
#view(tense_020526_sum)
M1imperfect <- glmmTMB(imperfect_per_100_words ~ Prompt + (1|Participant),
            	data = tense_020526_sum,
            	family = tweedie(link = "log"))
summary(M1imperfect)
r.squaredGLMM(M1imperfect)
#emmeans
emm_imp <- emmeans(M1imperfect, ~ Prompt, type = "response")
emm_imp
contrast(emm_imp, method = "revpairwise")  # differences on response scale

#effect size with emmeans package
emm <- emmeans(M1imperfect, "Prompt")
eff_size(emm, sigma = sigma(M1imperfect), edf = df.residual(M1imperfect))
```

#### Check assumptions

```{r, echo=FALSE}
# Check Simulated residuals- residual diagnostics
sim_res <- DHARMa::simulateResiduals(M1imperfect)
plot(sim_res)
#additional tests
DHARMa::testUniformity(sim_res)     # global model fit
DHARMa::testDispersion(sim_res)     # over/under-dispersion
DHARMa::testZeroInflation(sim_res)  # unexpected excess zeros
```

# P-value adjustment

```{r, echo=FALSE}
pvalues<-c(0.000222,
0.000305,
0.0042,
0.0162,
0.017,
0.0228,
0.045,
0.087,
0.275,
0.348,
0.4)

fdrs<-p.adjust(pvalues, method="BH")
print(fdrs)
```

