---
title: "Fluency Analysis (Prompt Valence)"
author: "Paloma Fernandez Mira"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)

# Core
library(tidyverse)
library(janitor)

# Mixed effects + inference
library(lme4)
library(lmerTest)   # p-values for lmer (Satterthwaite)
library(MuMIn)      # r.squaredGLMM()
library(emmeans)

# Diagnostics
library(lattice)      # qqmath()
library(sjPlot)       # plot_model()
library(car)          # durbinWatsonTest()
library(performance)  # check_singularity(), check_collinearity()
library(JWileymisc)   # modelDiagnostics()

# Tables
library(broom.mixed)
library(gt)

# Input file
STORY_CSV <- "fluency_story_level.csv"

# If story file contains multiple speakers, set this to the learner label.
LEARNER_SPEAKER <- "Speaker2"

# Toggle: full diagnostics for each model (can make the PDF long)
DO_DIAGNOSTICS <- FALSE
```

# Research question and operationalizations

We compare two prompts:

- **Negative valence:** "tell a story about a time you or someone you know felt in danger"
- **Positive valence:** "tell a story about a time you or someone you know had a good time"

Fluency is operationalized at story level:

1. **Speech rate**: lexical tokens per minute
   Learner speaking time = sum of learner-turn durations within the story (from timestamps).
   Lexical tokens include Spanish + English words, excluding obvious filled pauses.

2. **Breakdown fluency (without silent pauses)**:
   - Filled pause rate per minute
   - Filled pause rate per 100 lexical tokens

3. **L1 reliance (English rate)**:
   - English tokens per 100 lexical tokens

4. **Non-Spanish token rate**:
   - (filled pauses + English tokens) per 100 lexical tokens

5. **False starts**:
   - False starts per minute

# Load and prepare data

```{r load-data}
df <- read_csv(STORY_CSV, show_col_types = FALSE) |> clean_names()

# Keep learner speaker only (if multiple speakers exist)
if ("speaker" %in% names(df)) {
  df <- df |> filter(speaker == LEARNER_SPEAKER)
}

glimpse(df)
```

## Sample check

```{r sample-check}
df |>
  dplyr::count(course, prompt) |>
  arrange(course, prompt) |>
  gt()

df |>
  dplyr::summarise(
    n_rows         = n(),
    n_participants = n_distinct(participant_id),
    n_courses      = n_distinct(course)
  ) |>
  gt()
```

# Derive analysis variables

```{r derive-measures}
df <- df |>
  dplyr::mutate(
    participant_id = factor(participant_id),
    course         = factor(course),
    prompt         = factor(prompt),

    valence = case_when(
      str_to_lower(as.character(prompt)) %in% c("bad time","bad_time","negative","neg") ~ "Negative",
      str_to_lower(as.character(prompt)) %in% c("good time","good_time","positive","pos") ~ "Positive",
      TRUE ~ as.character(prompt)
    ),
    valence = factor(valence, levels = c("Negative", "Positive"))
  ) |>
  mutate(
    # 1) Speech rate
    speech_rate_tpm = coalesce(
      kept_lexical_tokens_per_min,
      total_kept_lexical_tokens_excl_false / total_minutes
    ),

    # 2) Filled pause rates
    fp_per_min    = coalesce(
      raw_filler_tokens_per_min,
      total_raw_filler_tokens / total_minutes
    ),
    fp_per_100lex = 100 * total_raw_filler_tokens / pmax(total_kept_lexical_tokens_excl_false, 1),

    # 3) English rate (L1 reliance)
    english_per_100lex = coalesce(
      removed_lexical_est_english_per_100_kept_lexical,
      100 * total_removed_lexical_est_english / pmax(total_kept_lexical_tokens_excl_false, 1)
    ),

    # 4) Non-Spanish token rate
    non_spanish_per_100lex = fp_per_100lex + english_per_100lex,

    # 5) False starts
    false_starts_per_min2   = coalesce(
      false_starts_per_min,
      total_false_start_tokens / total_minutes
    ),
    false_starts_per_100lex = 100 * total_false_start_tokens / pmax(total_kept_lexical_tokens_excl_false, 1)
  )

# Outcomes included in this analysis
dvs <- c(
  "speech_rate_tpm",
  "fp_per_min",
  "fp_per_100lex",
  "english_per_100lex",
  "non_spanish_per_100lex",
  "false_starts_per_min2"
)

# Bonferroni for this set
bonf_n     <- length(dvs)
bonf_alpha <- 0.05 / bonf_n
bonf_alpha
```

# Inspect distributions

```{r histograms, fig.height=7}
par(mfrow = c(3, 2))
hist(df$speech_rate_tpm,        main = "Speech rate (tokens/min)",             xlab = "tokens/min")
hist(df$fp_per_min,             main = "Filled pauses per minute",             xlab = "FP/min")
hist(df$fp_per_100lex,          main = "Filled pauses per 100 lexical tokens", xlab = "FP/100lex")
hist(df$english_per_100lex,     main = "English per 100 lexical tokens",       xlab = "Eng/100lex")
hist(df$non_spanish_per_100lex, main = "Non-Spanish per 100 lexical tokens",   xlab = "(FP+Eng)/100lex")
hist(df$false_starts_per_min2,  main = "False starts per minute",              xlab = "FS/min")
par(mfrow = c(1, 1))
```

# Linear mixed-effects models (LMER)

We fit one LMER per outcome:

- **Fixed effect:** prompt valence
- **Random intercepts:** course and student nested within course
  `DV ~ valence + (1 | course/participant_id)`

Bonferroni threshold for these `r bonf_n` outcomes: **`r signif(bonf_alpha, 5)`**.

```{r model-helpers}
fit_lmer <- function(data, dv) {
  f <- as.formula(paste0(dv, " ~ valence + (1|course/participant_id)"))
  lmer(
    f, data = data,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  )
}

summarize_model <- function(mod) {
  print(summary(mod))
  print(r.squaredGLMM(mod))
  print(emmeans(mod, specs = pairwise ~ valence))
  cat("\nSingularity check:\n")
  print(performance::check_singularity(mod))
}
```

## Run models

```{r run-models}
mods <- setNames(vector("list", length(dvs)), dvs)

for (dv in dvs) {
  cat("\n\n==============================\n")
  cat("MODEL:", dv, "\n")
  cat("==============================\n\n")
  mods[[dv]] <- fit_lmer(df, dv)
  summarize_model(mods[[dv]])
}
```

# Slide-style plots

```{r slide-plots, echo=FALSE, fig.height=5, fig.width=6}
# Human-readable labels for each DV
dv_labels <- c(
  speech_rate_tpm        = "Speech rate (lexical tokens/min)",
  fp_per_min             = "Filled pauses/min",
  fp_per_100lex          = "Filled pauses/100 lexical tokens",
  english_per_100lex     = "English/100 lexical tokens",
  non_spanish_per_100lex = "Non-Spanish/100 lexical tokens",
  false_starts_per_min2  = "False starts/min"
)

# Colour palette matching slide (green = Negative, orange = Positive)
valence_colors <- c("Negative" = "#2E8B6E", "Positive" = "#D4611A")

# Extract R2c for each model (mods already exists from run-models chunk)
r2c_values <- sapply(dvs, function(dv) {
  r2 <- MuMIn::r.squaredGLMM(mods[[dv]])
  unname(r2[, "R2c"])
})

plot_slide_style <- function(data, dv, ylab) {

  ggplot(data = data,
         aes(x = valence, y = .data[[dv]], color = valence)) +

    # Violin with median line drawn across full width via draw_quantiles
    geom_violin(draw_quantiles = 0.50, trim = FALSE, fill = NA, linewidth = 0.9) +

    # Jittered points
    geom_jitter(shape = 16, position = position_jitter(0.2), alpha = 0.8, size = 2) +

    scale_color_manual(values = valence_colors) +
    scale_x_discrete(labels  = c("Negative" = "bad_time", "Positive" = "good_time")) +

    labs(x = "Prompt", y = ylab, color = NULL) +

    theme_bw(base_size = 14) +
    theme(
      legend.position  = "none",
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      plot.background  = element_rect(fill = "white", color = NA),
      axis.title       = element_text(size = 13),
      axis.text        = element_text(color = "grey40", size = 12)
    )
}

# Print one plot per DV
for (dv in dvs) {
  p <- plot_slide_style(
    data    = df,
    dv      = dv,
    ylab    = dv_labels[[dv]]
  )
  print(p)
}
```

# Manuscript-ready results table (valence effect)

This table reports the **fixed effect of Positive vs Negative** prompt valence for each outcome,
including 95% CI, p-value, Bonferroni-adjusted p-value, and R2 (marginal/conditional).

```{r manuscript-table}
extract_valence_row <- function(mod, outcome) {
  te <- broom.mixed::tidy(mod, effects = "fixed", conf.int = TRUE) |>
    filter(term == "valencePositive") |>
    mutate(outcome = outcome)

  r2 <- MuMIn::r.squaredGLMM(mod)
  te$marginal_r2    <- unname(r2[1])
  te$conditional_r2 <- unname(r2[2])
  te
}

tab <- map_dfr(names(mods), ~ extract_valence_row(mods[[.x]], .x)) |>
  mutate(
    p_bonf   = p.adjust(p.value, method = "bonferroni", n = bonf_n),
    sig_bonf = if_else(!is.na(p.value) & p.value < bonf_alpha, "YES", "no")
  ) |>
  select(
    outcome,
    estimate, conf.low, conf.high, std.error, df, statistic, p.value, p_bonf,
    marginal_r2, conditional_r2, sig_bonf
  )

label_map <- c(
  speech_rate_tpm        = "Speech rate (tokens/min)",
  fp_per_min             = "Filled pauses/min",
  fp_per_100lex          = "Filled pauses/100 lexical",
  english_per_100lex     = "English/100 lexical",
  non_spanish_per_100lex = "Non-Spanish/100 lexical (FP+Eng)",
  false_starts_per_min2  = "False starts/min"
)

tab <- tab |>
  mutate(outcome = dplyr::recode(outcome, !!!label_map))

tab |>
  gt() |>
  tab_header(title = "Valence effect (Positive vs Negative) across fluency outcomes") |>
  fmt_number(
    columns  = c(estimate, conf.low, conf.high, std.error, statistic, p.value, p_bonf, marginal_r2, conditional_r2),
    decimals = 4
  ) |>
  cols_label(
    estimate       = "b",
    conf.low       = "CI low",
    conf.high      = "CI high",
    std.error      = "SE",
    df             = "df",
    statistic      = "t",
    p.value        = "p",
    p_bonf         = "p (Bonf)",
    marginal_r2    = "R2m",
    conditional_r2 = "R2c",
    sig_bonf       = paste0("p < ", signif(bonf_alpha, 5), "?")
  )
```

# Diagnostics (optional)

```{r diagnostics, eval=TRUE}
if (DO_DIAGNOSTICS) {
  for (dv in names(mods)) {
    mod <- mods[[dv]]
    cat("\n\n==============================\n")
    cat("DIAGNOSTICS:", dv, "\n")
    cat("==============================\n\n")

    print(sjPlot::plot_model(mod, type = "slope"))

    plot(fitted(mod), residuals(mod), main = paste("Residuals vs fitted:", dv))
    abline(h = 0, lty = 2)

    lattice::qqmath(mod, main = paste("QQ (random effects):", dv))
    qqnorm(residuals(mod), main = paste("QQ (residuals):", dv))
    qqline(residuals(mod))

    if (requireNamespace("glmmTMB", quietly = TRUE)) {
      print(sjPlot::plot_model(mod, type = "diag"))
    } else {
      cat("Skipping sjPlot diag panel (install glmmTMB to enable).\n")
    }

    acf(residuals(mod), main = paste("ACF residuals:", dv))
    print(car::durbinWatsonTest(residuals(mod)))
    print(performance::check_collinearity(mod))
    print(performance::check_model(mod))
    print(performance::check_normality(mod))
    print(performance::check_heteroscedasticity(mod))
    print(performance::check_outliers(mod))
  }
}
```

# Effect sizes

## Speech rate 

```{r eff_size, echo=FALSE}
#effect size
#view(df) speech_rate_tpm
#first have to pivot tibble wider
df_cd_speech_rate <- df|>
  dplyr::select(participant_id, prompt, speech_rate_tpm)|>
  pivot_wider(names_from = prompt, values_from = speech_rate_tpm)
#view(df_cd_speech_rate)

#Cohen's d
effectsize::cohens_d(df_cd_speech_rate$`bad time`, df_cd_speech_rate$`good time`, pooled_sd = TRUE, paired = TRUE)
#"Most of the time the actual population standard deviation is not known, this is why we estimate it using a pooled standard deviation from our two groups." (National University Stats Center)
```

## Filled pauses/100 lexical tokens


```{r eff_size1, echo=FALSE}
#first have to pivot tibble wider
df_cd_fp_per_100lex <- df|>
  dplyr::select(participant_id, prompt, fp_per_100lex)|>
  pivot_wider(names_from = prompt, values_from = fp_per_100lex)
#view(df_cd_fp_per_100lex)

#Cohen's d
effectsize::cohens_d(df_cd_fp_per_100lex$`bad time`, df_cd_fp_per_100lex$`good time`, pooled_sd = TRUE, paired = TRUE)
```